---
title: (Almost) every technique is a linear model
subtitle: "BIOL2022: **B**iology **E**xperimental **D**esign and **A**nalysis (**BEDA**)"
author:
  - name: Dr Januar Harianto
    affiliations: The University of Sydney
embed-resources: false # embed resources into single HTML file -- enable when ready to publish
# NOTE: please check _quarto.yml file for more options
format:
  ochre-revealjs: default
---


# The general linear model (GLM)

A *general (mathematical) framework* for expressing relationships between variables.


## The basic idea

 We are trying to fit a line (or a curve) to the data that best explains the relationship between the variables, essentially a simple equation many of you will be familiar with:

$$y = c + mx$$

where $c$ is the intercept and $m$ is the slope of the line.

::: fragment
Compare this to the general linear model:
$$y = \beta_0 + \beta_1 x + \epsilon$$
:::


## $$y = \beta_0 + \beta_1 x + \epsilon$$

or simply (for your benefit):

### $$y \sim x$$


::: fragment
**The equation to rule them all**, where

- $y$ is the response variable, or the variable we are trying to predict or explain.
- $x$ is a predictor variable that can be *anything*, continuous, categorical, binary, ordinal, etc.
- $\beta_i$, where $i$ is an integer, are the coefficients/estimates that quantify the relationship between $y$ and $x$. 
- $\epsilon$ is the error term, which captures the variability in $y$ that is not explained by $x$.
:::

::: fragment
**In general, most common statistical tests can be expressed as a GLM like above.**
:::

## $$y = \beta_0 + \beta_1 x + \epsilon$$
### Why does it matter?

::: incremental
 - Under the GLM, the statistical test is identified based on the variables supplied to the model. This follows a model-centric approach:
  - If the predictor variable ($x$) is **binary**, we are performing a GLM *equivalent* to a ***t*-test** and the slope is the difference in conditional means.
  - If the predictor variable ($x$) is **categorical**, we are performing a GLM *equivalent* to **ANOVA**, and the slope represents the difference in means between the categories.
  - If the predictor variable ($x$) is **continuous**, we are performing a GLM *equivalent* to **regression**. The slope becomes the difference in expected values for a pair of points that differ in $x$ by one unit.
    - *For one unit change in $x$, the expected change in $y$ is $\beta_1$.*
- The GLM takes care of the "statistical test" for you and you *do not* need to know what the equivalent test is (although it helps).

:::

## How we can use it

::: fragment
1. Model the relationship between the response variable and the predictor/explanatory variable(s) as a simple function.

:::
::: fragment
$$response \sim explanation$$

Also read as (recall the previous lecture):

- response is explained by the explanation
- response is a function of the explanation
- response is influenced by the explanation
- ...
:::

## How we can use it {auto-animate=true}

1. Model the relationship between the response variable and the predictor/explanatory variable(s) as a simple function.

$$response \sim explanation$$

2. Define the **structure** of the variables in the model, which can be continuous, categorical, binary, etc., something we can decide if we are the ones designing the study.

::: fragment
$$ (continuous) \sim (continuous) $$

$$ (continuous) \sim (categorical) $$

$$ (binary) \sim (continuous) $$
:::

## How we can use it {auto-animate=true}

1. Model the relationship between the response variable and the predictor/explanatory variable(s) as a simple function.

$$response \sim explanation$$

2. Define the **structure** of the variables in the model, which can be continuous, categorical, binary, etc., something we can decide if we are the ones designing the study.

$$ (continuous) \sim (categorical) $$

::: fragment
3. Fit the model to the data based on the data types, and start work out if the model is a good fit.

$$ response = \beta_0 + \beta_1 \times explanation + \epsilon $$

:::



## How we can use it {auto-animate=true}
$$response \sim explanation$$

$$ (continuous) \sim (categorical) $$

$$ response = \beta_0 + \beta_1 \times explanation + \epsilon $$

::: {.fragment}
In most statistical software (including R and Jamovi) you only need to specify the model formula (the first part) and the software will take care of the rest to produce the statistical output (the third part).

**Your job is to determine if the model is a good fit, and interpret the results.**
:::

# A good fit?

## Checking assumptions: many ways

In previous studies, you may have checked assumptions in different ways depending on the statistical test:

- For a *t*-test, you may have checked for normality by plotting the data and looking for a bell-shaped curve. Equal variances you may have used the F-test or Levene's test.
- For ANOVA, you may have checked for homogeneity of variance using Levene's test or Bartlett's test.
- For regression, you looked at the residuals to check for homoscedasticity and normality.

## Checking assumptions: the GLM way

A GLM provides a framework for checking model assumptions using residuals to assess any of:

- **Homoscedasticity** and **Linearity**: Are the residuals spread evenly across the range of the predictor variable?
- **Normality**: Are the residuals normally distributed?
- **Outliers**: Are there any outliers that are affecting the model?
- **Influential points**: Are there any points that are affecting the model?
- **Collinearity**: Are the predictor variables correlated with each other?

We will cover these in more detail in the next lecture...



## Preview: continuous predictor

```{r}
#| echo: false
fit <- lm(mpg ~ wt, data = mtcars)
par(mfrow = c(2, 2))
plot(fit)
```

## Preview: categorical predictor

```{r}
#| echo: false

# not for anova equivalent using chick data
fit <- lm(weight ~ feed, data = chickwts)
par(mfrow = c(2, 2))
plot(fit)
```

# Expanding the GLM to include more variables

## Expanding the GLM: More than one variable {auto-animate=true}

So far, we have looked at models with a single explanatory variable. But often, we want to explain the response variable using **more than one** explanatory variable.

For example, perhaps we want to explain the metabolic rate of an organism using both its body mass **and** temperature:

$$\text{metabolic rate} \sim \text{body mass} + \text{temperature}$$

::: fragment
### What does the plus sign mean?

The plus sign (`+`) means we are including **both** body mass and temperature as separate, additive predictors of metabolic rate. This is called an **additive model**.
:::

## Adding interactions {auto-animate=true}

$$\text{metabolic rate} \sim \text{body mass} + \text{temperature}$$

Sometimes, the effect of one variable depends on the value of another. For example, maybe the effect of temperature on metabolic rate changes depending on body mass. 

::: {.fragment}
To model this, we include an **interaction**:

$$\text{metabolic rate} \sim \text{body mass} \times \text{temperature}$$
:::

::: {.fragment}

This expands to:

$$\text{metabolic rate} \sim \text{body mass} + \text{temperature} + \text{body mass}\mathbin{:}\text{temperature}$$

- The `:` term represents the **interaction** between body mass and temperature.
- The `*` shorthand (`body mass * temperature`) automatically includes both main effects and their interaction.

There are other types of relationships and model structures, which we will cover as we go along.
:::

## Quick exercises

Model the following relationships.

::: fragment
- The number of birds in a forest is influenced by the number of trees and the amount of rainfall.
:::

::: fragment
$$ \text{number of birds} \sim \text{number of trees} + \text{amount of rainfall} $$
:::

::: fragment
- The number of birds in a forest is influenced by the number of trees and the amount of rainfall, and there is an interaction between the number of trees and the amount of rainfall.
:::

::: fragment
$$ \text{number of birds} \sim \text{number of trees} + \text{amount of rainfall} + \text{number of trees}\mathbin{:}\text{amount of rainfall} $$

<center>or</center>

$$ \text{number of birds} \sim \text{number of trees} \times \text{amount of rainfall} $$
:::

## Quick exercises

Model the following relationships.

::: fragment
- Whether a turtle is born male or female is influenced by the temperature of the nest.
:::

::: fragment
$$ \text{sex of turtle} \sim \text{temperature of nest} $$
:::

::: fragment
- The height of a seedling is influenced by temperature and the amount of sunlight it receives.
:::

::: fragment
$$ \text{height of seedling} \sim \text{temperature} + \text{amount of sunlight} $$
:::

::: fragment
- Air quality is influenced by rainfall, temperature, the number of trees and cloud cover, and there is an interaction between rainfall and temperature.
:::

::: fragment
$$ \text{air quality} \sim \text{rainfall} + \text{temperature} + \text{number of trees} + \text{cloud cover} + \text{rainfall}\mathbin{:}\text{temperature} $$
:::




## So how do I use it as a newbie?

You are probably **NOT experienced enough to use it to its full potential**, and that is okay.

::: fragment
- Use the GLM as a **general framework to model relationships** between variables.
- What you *can* do is:
  1. Define (model) the relationship between the response and explanatory variables.
  2. Identify the **statistical test** the GLM is *related* to.
  3. Run the statistical test, but use GLM **as a guide** to diagnose (e.g. assumptions) and interpret the results.
:::


# (If there is time) Examples in Jamovi and R





# Thanks!

This presentation is based on the [SOLES Quarto reveal.js template](https://github.com/usyd-soles-edu/soles-revealjs) and is licensed under a [Creative Commons Attribution 4.0 International License][cc-by].


<!-- Links -->
[cc-by]: http://creativecommons.org/licenses/by/4.0/